Updates a Procrustes and a PCA

### 2025-04-14 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@

The function that does it automatically is now implemented in `landvR`.

@@

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, individuals_array, scale.pca = FALSE)

## BUG IN THE NUMBER OF DIMENSIONS!
dim(updated_pca)[1] == dim(whole_gpa$PCscores)[1] + length(all_individuals)
dim(updated_pca)[2] == dim(whole_gpa$PCscores)[2]

## SOME SMALL SHIFT!
plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
## Plot the chaffinches to see where they lay:
fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
```

>>>>

```{r}
library(Morpho)
library(landvR) # install from devtools::install_github("TGuillerme/landvR")

@@

But does it work with the same specimens? I.e. does it place previous specimens in the correct position?

```{r}
## Old specimens
old_spec <- array(data = c(original_landmarks[,,1],
                           original_landmarks[,,2],
                           original_landmarks[,,3]), dim = c(10, 3, 3))
dimnames(old_spec)[[3]] <- paste0("old_", c("spec1", "spec2", "spec3"))

## Updating the gpa and pca
updated_pca <- update.gpa.pca(proc_orig, old_spec, center.pca = TRUE, scale.pca = FALSE)

## Plot all the results
plot(proc_orig$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5, col = "grey")

points(proc_orig$PCscores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
points(updated_pca[c(81:83),c(1,2)], col = "orange", pch = 19)

legend("topright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "old_spec (target)", "old_spec (results)"))
```

This is not ideal!
The reason why it doesn't recover them is because it slightly offsets the superimposition which cascades to big changes in the PCA.

For example, when using the same specimens:

```{r}
## Plotting the landmarks before superimposition
op <- par(mfrow = c(2,2))
plot(old_spec[, c(1,2), 1], main = "2D landmarks for spec 1", xlab = "", ylab = "")
plot(old_spec[, c(1,2), 2], main = "2D landmarks for spec 2", xlab = "", ylab = "")
plot(old_spec[, c(1,2), 3], main = "2D landmarks for spec 3", xlab = "", ylab = "")
par(op)

## Project the new specimens on the original GPA
new_gpa <- Morpho::align2procSym(proc_orig, old_spec, orp = TRUE)

## Slightly off?
op <- par(mfrow = c(2,2))
## Spec 1
plot.offset <- function(id, new_gpa, old_gpa) {
    main <- paste0("spec ", id, " reprojected onto itself\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
}

plot.offset(id = 1, new_gpa, proc_orig$rotated)
plot.offset(id = 2, new_gpa, proc_orig$rotated)
plot.offset(id = 3, new_gpa, proc_orig$rotated)

main <- paste0("spec 1 projected onto spec 2 (for ref.)\nTotal offset = ", sum(sqrt((proc_orig$rotated[, 1, 1] - proc_orig$rotated[, 1, 2])^2 + (proc_orig$rotated[, 2, 1] - proc_orig$rotated[, 2, 2])^2)))
plot(proc_orig$rotated[, c(1,2), 1], main = main, xlab = "", ylab = "")
points(proc_orig$rotated[, c(1,2), 2], pch = 19, cex = 0.5)

par(op)
```

This behaviour is actually highlighted in the notes for `align2procSym`: "this will never yield the same result as a pooled Procrustes analysis because the sample mean is iteratively updated and new data would change the mean."

> TL;DR: it's not working until we find a way to not update the mean.

@@

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, whole_gpa$rotated[,,c(2819, 2820)], scale.pca = FALSE)

## Some big shift!
plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
## Plot the chaffinches to see where they lay:
fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
```

## Creating some new specimens (two old ones, for benchmarking the method)
new_spec <- array(c(original_landmarks[,,1],
                    original_landmarks[,,2]), dim = c(10, 3, 2))

## Project the new specimens on the original GPA
new_gpa <- Morpho::align2procSym(proc_orig, new_spec, orp = TRUE)

## Slightly off?
plot.offset <- function(id, new_gpa, old_gpa, main) {
    main <- paste0(main, "\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
}

op <- par(mfrow = c(2,1))
plot.offset(id = 1, new_gpa, proc_orig$rotated, main = "Spec 1 projected onto itself")
plot.offset(id = 2, new_gpa, proc_orig$rotated, main = "Spec 2 projected onto itself")
par(op)

data(boneData)
# run procSym on entire data set
proc <- procSym(boneLM)
# this is the training data
array1 <- boneLM[,,1:60]
proc1 <- procSym(array1)
newdata <- boneLM[,,1:2]
newalign <- align2procSym(x = proc1, newdata = newdata)

plot.offset(id = 1, newalign, proc1$rotated, main = "Spec 1 projected onto itself")

```

TODO: try doing the alignement of some species on the edges of the traitspace and see if it falls back on the edges or if it goes to the middle again.

TODO: check with the `morphospace` package?
